---
title: "Figure 2 and S3: Adnp2 heatmap and repeat element overlap"
author: Michaela Schwaiger
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(width = 90L)
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(MiniChip)
  library(GenomicRanges)
  library(ComplexHeatmap)
  library(circlize)
  library(cowplot)
  library(UpSetR)

})

#parameters
topdir <- "../"
figuredir <-  "figures"
datadir <- "data"


span = 1500
step = 15
```

# Figure 2A: heatmap of Adnp2, Chd4, and HP1b

```{r, fig.width=10,fig.height=10}
SEreadShift <- 50

counts <- readRDS(file.path(topdir,datadir,sprintf("heatmap_counts_at_Adnp_and_Adnp2_peaks_SEreadShift_%d.rds",SEreadShift)))
#select samples 
HMcounts1 <- counts[c("Adnp2_WT_ut","Chd4","HP1b")]
#select peaks
HMcounts <- lapply(HMcounts1,function(x){x[grep("Adnp2_",rownames(HMcounts1[[1]])),]})

medianCpm <- c(0.04,0.02,0.02)
#medianCpm <- c(1.5,0.7,0.7)
topCpm=c(100,100,100)

plotcols <- c("#6e377d","#008ca5","#ebb400")
  
#plot heatmaps
heatlist <- DrawSummitHeatmaps(
  HMcounts,
  bamNames = names(HMcounts),
  plotcols = plotcols,
  use.log = FALSE,
  medianCpm = medianCpm,
  topCpm = topCpm,
  TargetHeight = 500,
  orderSample = 1,
  orderWindows = 15,
  summarizing = "mean",
  show_axis = FALSE,
  MetaScale = c(rep("all", length(HMcounts)))
)

#pdf(file.path(topdir,figuredir,sprintf("Figure2A_Adnp2_Chd4_HP1b_heatmaps_at_Adnp2_peaks_%d_SEreadShift.pdf",SEreadShift)), width=4,height=7)
draw(heatlist, padding = unit(c(3, 8, 8, 2), "mm"),show_heatmap_legend=TRUE)
#dev.off()
```

# Figure 2B:  overlap of Adnp2 peaks with annotations - upsetplot

```{r, fig.width=10,fig.height=7}
ADNP2_4_upset <- read.table(file=file.path(topdir,datadir,"Adnp2_peaks_for_upset_plot.txt"),sep="\t",header=TRUE)
upset(ADNP2_4_upset, nsets = 5, nintersects = 10, set_size.show = TRUE, order.by= "freq")
```

# Figure 2C:  overlap of Adnp2 peaks with different repeat classes - barplot

```{r}

repoverlap <- read.table(file.path(topdir,datadir,"repeat_names_andTSS_at_Adnp2_vs_random_peaks.txt"),sep="\t",header=TRUE)
repoverlap$ChIP <- "ADNP2"
colnames(repoverlap) <- c("repeat_name", "peak", "random", "sd",  "repeat_class","ChIP")
repoverlap$label <- ifelse(repoverlap$ChIP=="ADNP2" & repoverlap$peak > 150 & repoverlap$random < 200,repoverlap$repeat_name,
                          ifelse(repoverlap$ChIP=="ADNP" & repoverlap$peak > 1700 & repoverlap$random < 800,repoverlap$repeat_name,"" ))
repoverlap$repclass <- ifelse(grepl("^TSS",repoverlap$repeat_class),"TSS",
                              ifelse(grepl("^DNA",repoverlap$repeat_class),"DNA",
                              ifelse(grepl("RNA",repoverlap$repeat_class),"RNA",
                                     ifelse(grepl("^LINE",repoverlap$repeat_class),"LINE",
                                         ifelse(grepl("^LTR",repoverlap$repeat_class),"LTR",
                                                ifelse(grepl("^SINE",repoverlap$repeat_class),"SINE",
                                                       ifelse(grepl("Satellite",repoverlap$repeat_class),"Satellite",
                                                              ifelse(repoverlap$repeat_class=="Simple_repeat" | repoverlap$repeat_class=="Low_complexity","Simple/low_complexity","other"))))))))

#summarize the repeat overlaps by repeat class
repoverlap2 <- repoverlap %>% group_by(repclass) %>% summarize(peak=sum(peak),random=sum(random)) %>% pivot_longer(cols = c("peak", "random"),names_to = "peaktype",values_to = "number_of_repeats")

#sort by number of peaks overlapping
repoverlap3 <- repoverlap2[repoverlap2$peaktype=="peak",]
repoverlap3 <- repoverlap3[order(repoverlap3$number_of_repeats,decreasing=TRUE),]
repoverlap2$repclass <- factor(repoverlap2$repclass,levels=repoverlap3$repclass,labels=repoverlap3$repclass)

#plot
ggplot(repoverlap2[repoverlap2$repclass %in% c("LTR", "LINE", "TSS", "SINE", "Simple/low_complexity"),],aes(y=number_of_repeats,x=repclass,fill=peaktype)) + geom_bar(stat="identity",position = "dodge") + 
  theme_classic() + scale_fill_manual(values = c("#6e377d","grey")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
#ggsave(filename=file.path(topdir,figuredir,"Figure2Ba_repeat_names_andTSS_at_Adnp2_peaks_barplot.pdf"),height=4, width=6,device = "pdf")
```

# Figure 2D: detailed overlap of Adnp2 peaks with different repeat elements

```{r}

repoverlap <- read.table(file.path(topdir,datadir,"repeat_names_andTSS_at_Adnp2_vs_random_peaks.txt"),sep="\t",header=TRUE)
repoverlap$ChIP <- "ADNP2"
colnames(repoverlap) <- c("repeat_name", "peak", "random", "sd",  "repeat_class","ChIP")
repoverlap$label <- ifelse(repoverlap$ChIP=="ADNP2" & repoverlap$peak > 150 & repoverlap$random < 200,repoverlap$repeat_name,
                          ifelse(repoverlap$ChIP=="ADNP" & repoverlap$peak > 1700 & repoverlap$random < 800,repoverlap$repeat_name,"" ))
repoverlap$repclass <- ifelse(grepl("^TSS",repoverlap$repeat_class),"TSS",
                              ifelse(grepl("^DNA",repoverlap$repeat_class),"DNA",
                              ifelse(grepl("RNA",repoverlap$repeat_class),"RNA",
                                     ifelse(grepl("^LINE",repoverlap$repeat_class),"LINE",
                                         ifelse(grepl("^LTR",repoverlap$repeat_class),"LTR",
                                                ifelse(grepl("^SINE",repoverlap$repeat_class),"SINE",
                                                       ifelse(grepl("Satellite",repoverlap$repeat_class),"Satellite",
                                                              ifelse(repoverlap$repeat_class=="Simple_repeat" | repoverlap$repeat_class=="Low_complexity","Simple/low_complexity","other"))))))))

#plot
ggplot(repoverlap,aes(y=peak,x=random,col=repclass)) + geom_point() + theme_classic() +
 # geom_text(aes(label=ifelse(peak > 200 & random < 50,as.character(repeat_name),'')),hjust=0.5,vjust=-0.7,size=2) +
  geom_text(aes(label=label),hjust=0.1,vjust=-0.5,size=4) +
  #xlim(0,1000) + ylim(0,1000) + 
  # scale_color_discrete()  + 
  geom_errorbarh(aes(xmax = random + sd, xmin = random - sd, height = 5)) +
  geom_abline(slope = 1, intercept = 0,col="grey",linetype="dotted") +
  facet_wrap(vars(ChIP),scales="free") + scale_color_manual(values = c("TSS" ="white" ,
                                "DNA"="#E27069FF","RNA"="#DBA520FF","LINE"="#E93FBCFF","LTR"="#2E8B57FF",
                               "SINE"="#1175BBFF","Satellite"="#7852A9FF","Simple/low_complexity"="#4AC6AEFF","other"= "#EF7215FF"))

#ggsave(filename=file.path(topdir,figuredir,"Figure2B_repeat_names_andTSS_at_Adnp2_vs_random_peaks.pdf"),height=4, width=6,device = "pdf")
```

# Figure 2E and S3A: repeat cpm barplots

```{r,fig.height=5,fig.width=10}
cpms2sumsAves <- readRDS(file.path(topdir,datadir,"per_repeat_family_per_group_mean_cpms.rds"))
cpms2sumsSDs <- readRDS(file.path(topdir,datadir,"per_repeat_family_per_group_sd_cpms.rds"))

cpms2sumsAves_long <- cpms2sumsAves[,1:16] %>% pivot_longer(cols=colnames(cpms2sumsAves[,1:16])[4:ncol(cpms2sumsAves[,1:16])],names_to = "sample",values_to="cpm")
cpms2sumsSDs_long <- cpms2sumsSDs %>% pivot_longer(cols=colnames(cpms2sumsSDs)[4:ncol(cpms2sumsSDs)],names_to = "sample",values_to="cpm_sd")
cpms2sumsAves_long <- cbind(cpms2sumsAves_long,cpms2sumsSDs_long[,5])

cpms2sumsAves_long_sel <- cpms2sumsAves_long[cpms2sumsAves_long$repeat_name %in% c("IAPEz-int","MMERVK10C-int","RLTR10C","IAPLTR1a_Mm","L1MdA_I"),]
cpms2sumsAves_long_sel$ChIP <- ifelse(grepl("^H3K9me3",cpms2sumsAves_long_sel$sample),"H3K9me3",
                                      ifelse(grepl("^Adnp2",cpms2sumsAves_long_sel$sample),"ADNP2","ADNP"))
  
##########individual replicate data for dots##############
cpms2 <- readRDS(file.path(topdir,datadir,"per_repeat_per_sample_cpms.rds"))

cpms2sums <- cpms2 %>% group_by(repclass,repeat_class,repeat_name) %>% summarize_if(is.numeric, sum, na.rm = TRUE)
cpms3sums_long  <- cpms2sums[cpms2sums$repeat_name %in% c("IAPEz-int","MMERVK10C-int","RLTR10C","IAPLTR1a_Mm","L1MdA_I"),] %>% pivot_longer(cols=colnames(cpms2sums)[4:ncol(cpms2sums)],names_to = "sample",values_to="cpm")

#remove the extra (non-matched) replicates from ADNP2 data
#cpms3sums_long <- cpms3sums_long[grepl("_r[0-2]",cpms3sums_long$sample),]
cpms3sums_long$group <- gsub("_r[0-9]","",cpms3sums_long$sample)
cpms3sums_long$group <- ifelse(cpms3sums_long$group=="InputK9_WT_ut","H3K9me3_Input",
                               ifelse(cpms3sums_long$group=="Input_WT_ut","Adnp2_Input",
                                      ifelse(cpms3sums_long$group=="InputAdnp_WT","Adnp_Input",cpms3sums_long$group)))
                                             
cpms3sums_long$ChIP <- ifelse(grepl("^H3K9me3",cpms3sums_long$group),"H3K9me3",
                                      ifelse(grepl("^Adnp2",cpms3sums_long$group),"ADNP2",
                                        ifelse(grepl("^Adnp",cpms3sums_long$group),"ADNP",     "other")))
cpms3sums_long$replicate <- ifelse(grepl("_r1",cpms3sums_long$sample),"r1",
                                   ifelse(grepl("_r2",cpms3sums_long$sample),"r2",
                                          ifelse(grepl("_r3",cpms3sums_long$sample),"r3","r4")))
                                    

###ADNP2 Input and WT only
cpms2sumsAves_long_sel$group <- cpms2sumsAves_long_sel$sample
ggplot(cpms2sumsAves_long_sel[cpms2sumsAves_long_sel$group %in% c("Adnp2_Input","Adnp2_WT_ut"),],aes(x=group,y=cpm,fill=group)) + geom_bar(stat="identity") + facet_wrap(vars(repeat_name),scales="free",ncol=5) + theme_bw()  + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +  scale_fill_manual(values=c("grey","#6e377d")) +
  geom_errorbar(aes(ymin=cpm-cpm_sd, ymax=cpm+cpm_sd), width=.2,position=position_dodge(.9)) + 
  geom_jitter(data=cpms3sums_long[cpms3sums_long$sample %in% c("Input_WT_ut_r1","Input_WT_ut_r2", "Input_WT_ut_r3","Input_WT_ut_r4","Adnp2_WT_ut_r1","Adnp2_WT_ut_r2", "Adnp2_WT_ut_r3","Adnp2_WT_ut_r4"),],aes(x=group,y=cpm),col="darkgrey",position = position_jitter(0.1))
#ggsave(file.path(topdir,plotdir,"Adnp2_WT_repeat_cpm_ChIP_Vs_Input_selected_barplots_4reps_sd.pdf"),height=9,width=10)
```


# Figure 2F and S3B repeat consensus Adnp2 ChIP and Input coverage plots

```{r,fig.height=10,fig.width=8}
#load the coverage data
covTracks <- read.table(file=file.path(topdir,datadir,"repeat_consensus_Adnp2_coverage_tracks.txt"),sep="\t",header=TRUE,comment.char="^")

#select the L1MdA_I consensus sequences and combine them
covTracks1 <- filter(covTracks,repeat_name %in% c("L1MdA_I_5end#LINE/L1","L1MdA_I_orf2#LINE/L1","L1MdA_I_3end#LINE/L1"))

covTracks1$position2 <- ifelse(covTracks1$repeat_name=="L1MdA_I_orf2#LINE/L1",covTracks1$position + max(covTracks1$position[covTracks1$repeat_name=="L1MdA_I_5end#LINE/L1"]),
                               ifelse(covTracks1$repeat_name=="L1MdA_I_3end#LINE/L1",covTracks1$position + max(covTracks1$position[covTracks1$repeat_name=="L1MdA_I_5end#LINE/L1"]) + max(covTracks1$position[covTracks1$repeat_name=="L1MdA_I_orf2#LINE/L1"]),covTracks1$position))
covTracks1$group <- gsub("_r[1-4]","",covTracks1$sample)

L1plotGroup <- ggplot(covTracks1,aes(x=position2,y=cpm,col=group)) + geom_smooth(method = "loess",span=0.05) + ylab("cpm") + xlab("position on L1MdA_I") + 
  theme_classic() + scale_color_manual(values=c("#6a3d9a","darkgrey"))

#select the IAPEz consensus sequences and combine them
covTracks2 <- filter(covTracks,repeat_name %in% c("IAPEz-int#LTR/ERVK","IAPLTR1a_Mm#LTR/ERVK"))

#add internal sequence after LTR sequence
covTracks2$position2 <- ifelse(covTracks2$repeat_name=="IAPEz-int#LTR/ERVK",covTracks2$position + max(covTracks2$position[covTracks2$repeat_name=="IAPLTR1a_Mm#LTR/ERVK"]),
                               covTracks2$position)
#add inverted LTR sequence after internal sequence
covTracksLTR2 <- filter(covTracks,repeat_name %in% c("IAPLTR1a_Mm#LTR/ERVK"))
covTracksLTR2$position2 <- abs(covTracksLTR2$position - 338)
covTracksLTR2$position2 <- covTracksLTR2$position2 + max(covTracks2$position2)
covTracks2 <- rbind(covTracks2,covTracksLTR2)
covTracks2$group <- gsub("_r[1-4]","",covTracks2$sample)


IAPplotGroup <- ggplot(covTracks2,aes(x=position2,y=cpm,col=group)) + geom_smooth(method = "loess",span=0.05) + ylab("cpm") + xlab("position on IAPEz") + 
  theme_classic() + scale_color_manual(values=c("#6a3d9a","darkgrey"))
#select the MMERVK10C consensus sequences and combine them
covTracks3 <- filter(covTracks,repeat_name %in% c("MMERVK10C#LTR/ERVK","RLTR10C#LTR/ERVK"))

#add internal sequence after LTR sequence
covTracks3$position2 <- ifelse(covTracks3$repeat_name=="MMERVK10C#LTR/ERVK",covTracks3$position + max(covTracks3$position[covTracks3$repeat_name=="RLTR10C#LTR/ERVK"]),
                               covTracks3$position)
#add inverted LTR sequence after internal sequence
covTracksLTR3 <- filter(covTracks,repeat_name %in% c("RLTR10C#LTR/ERVK"))
covTracksLTR3$position2 <- abs(covTracksLTR3$position - 435)
covTracksLTR3$position2 <- covTracksLTR3$position2 + max(covTracks3$position2)
covTracks3 <- rbind(covTracks3,covTracksLTR3)
covTracks3$group <- gsub("_r[1-4]","",covTracks3$sample)


MMplotGroup <- ggplot(covTracks3,aes(x=position2,y=cpm,col=group)) + geom_smooth(method = "loess",span=0.05) + ylab("cpm") + xlab("position on MMERVK10C") + 
  theme_classic() + scale_color_manual(values=c("#6a3d9a","darkgrey"))

plot_grid(IAPplotGroup,L1plotGroup,MMplotGroup,nrow=3)
#ggsave(file=file.path(topdir,figuredir,"Figure2_repeat_consensus_Adnp2_coverage_reps_averaged.pdf"),height=9,width=6)
```


# Figure S3A:  overlap of Adnp2 peaks with different repeat classes - barplot

```{r}

repoverlap <- read.table(file.path(topdir,datadir,"repeat_names_andTSS_at_Adnp2_vs_random_peaks.txt"),sep="\t",header=TRUE)
repoverlap$ChIP <- "ADNP2"
colnames(repoverlap) <- c("repeat_name", "peak", "random", "sd",  "repeat_class","ChIP")
repoverlap$label <- ifelse(repoverlap$ChIP=="ADNP2" & repoverlap$peak > 50 & repoverlap$random < 200,repoverlap$repeat_name,
                          ifelse(repoverlap$ChIP=="ADNP" & repoverlap$peak > 1700 & repoverlap$random < 800,repoverlap$repeat_name,"" ))
repoverlap$repclass <- ifelse(grepl("^TSS",repoverlap$repeat_class),"TSS",
                              ifelse(grepl("^DNA",repoverlap$repeat_class),"DNA",
                              ifelse(grepl("RNA",repoverlap$repeat_class),"RNA",
                                     ifelse(grepl("^LINE",repoverlap$repeat_class),"LINE",
                                         ifelse(grepl("^LTR",repoverlap$repeat_class),"LTR",
                                                ifelse(grepl("^SINE",repoverlap$repeat_class),"SINE",
                                                       ifelse(grepl("Satellite",repoverlap$repeat_class),"Satellite",
                                                              ifelse(repoverlap$repeat_class=="Simple_repeat" | repoverlap$repeat_class=="Low_complexity","Simple/low_complexity","other"))))))))
table(repoverlap$repclass)

#summarize the repeat overlaps by repeat class
repoverlap2 <- repoverlap %>% group_by(label) %>% summarize(peak=sum(peak),random=sum(random)) %>% pivot_longer(cols = c("peak", "random"),names_to = "peaktype",values_to = "number_of_repeats")

#sort by number of peaks overlapping
repoverlap3 <- repoverlap2[repoverlap2$peaktype=="peak",]
repoverlap3 <- repoverlap3[order(repoverlap3$number_of_repeats,decreasing=TRUE),]
repoverlap2$label <- factor(repoverlap2$label,levels=repoverlap3$label,labels=repoverlap3$label)
#Drop unlabeled category
repoverlap2 <- repoverlap2[repoverlap2$label != "",]

#plot
ggplot(repoverlap2,aes(y=number_of_repeats,x=label,fill=peaktype)) + geom_bar(stat="identity",position = "dodge") + 
  theme_classic() + scale_fill_manual(values = c("#6e377d","grey")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave(filename=file.path(topdir,figuredir,"FigureS2A_repeat_names_andTSS_at_Adnp2_peaks_detailed_barplot.pdf"),height=4, width=6,device = "pdf")


```




